Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-07-08 paper/main.tex
minor change
fixed a typo
fixed rule Q:Return
removed preprint option
typo

% Comment for anonymization
% Also look for "SHORT" comments
% \def\fullversion{}

\documentclass[nocopyrightspace,preprint]{sigplanconf}

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice:
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that \iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
%
\itemskip
\begin{enumerate}
\item We define a operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

@@

@@

{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }

@@

@@

{ B; (\Gamma,Q_R); (\Gamma, Q_R{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }

@@

@@

We are only aware of two verified quantitative analysis systems.
Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
automatically verify previously inferred loop invariants, size
relations, and ranking functions for Java Card programs.  However,
they do not have a formal cost semantics and do not prove the bounds
correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
have verified a loop bound analysis for CompCert's RTL intermediate
language.  However, there is no way to interactively derive bounds or
to deal with resources like memory usage.  Blazy et al.'s automatic
bound analysis does also not compute symbolic bounds.

>>>>

% Comment for anonymization
% Add preprint option for the extended version
% Also look for "SHORT" comments
% \def\fullversion{}

\documentclass[nocopyrightspace]{sigplanconf}

@@

Static analysis of quantitative properties of imperative programs is
an active research area and recent years have seen many innovations.
Notable tools that have been developed include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.
%
While the these tools can derive impressive results for realistic
software, there still exist shortcomings that hamper the application
of existing techniques in practice:
\itemskip
\begin{itemize}
\item Analysis tools are black boxes that\iffull{either}{} deliver a result or fail
  without enabling \emph{user interaction} to manually or
  semi-automatically derive bounds for challenging parts of the
  program.
\itemskipIn
\item The computed bounds are often \emph{non-compositional} local
  bounds (for a single (nested) loop) that are difficult to combine to
  global whole program bounds.
\itemskipIn
\item Existing techniques often rely on complex external tools such as
  abstract interpretation--based invariant
  generation~\cite{GulwaniMC09} or translation of the program into a
  term-rewriting system~\cite{BrockschmidtEFFG14, SinnZV14} without
  providing \emph{verifiable certificates} for the correctness of the
  derived bound.
\end{itemize}
\itemskip
%
While there has been much progress in static quantitative analysis,
Knuth correctly points out in a recent interview~\cite{KnuthInter}
that the state-of-the-art in formal quantitative methods still falls
short in comparison with semantic techniques.
%
\itemskip
\begin{quote}
  % Consider, for example, a computer program that sorts a list of
  % numbers into order. 
  $[\cdots]$ Thanks to the work of Floyd, Hoare, and others,
  we have formal definitions of semantics, and tools by which we can
  verify that sorting is indeed always achieved. My job is to go
  beyond correctness, to an analysis of such things as the program's
  running time $[\cdots]$. I'm 100\% sure that my recurrence correctly
  describes the program's performance, and all of my colleagues agree
  with me that the recurrence is "obviously" valid. Yet I have no
  formal tools by which I can prove that my recurrence is right. I
  don't really understand my reasoning processes at all!
  \vspace{-4ex}
  \begin{flushright}
    -- Donald E. Knuth, 2014
  \end{flushright}
\end{quote}
\itemskip
%
In this work, we develop a resource analysis framework for C programs
that is based on a solid semantic foundation.  The choice of C is
primarily motivated by \iffull{our}{the} ongoing work on the formally verified
hypervisor kernel CertiKOS~\cite{GuVFSC11}.  CertiKOS is mainly
developed in C and is supposed to provide verified guaranties on
timing and memory usage.  Moreover, C is a natural choice because it
is still the most widely used language for system development; in
particular in safety-critical embedded and real-time systems where
resource bound analyses are often required by regulatory
authorities~\cite{May2013}.  Our \emph{contributions} are as follows:
%
\itemskip
\begin{enumerate}
\item We define an operational cost semantics for CompCert Clight that
  defines the resource consumption of terminating and diverging
  executions.  The cost is parametric in a user-definable cost metric
  and can be negative if resources are released during an execution.

@@

\\ \Gamma = \Gamma_R[\Vret/x]

@@

{ B; (\Gamma_R,Q_R); (\Gamma, Q) \vdash \code{return}~x \dashv (\Gamma',Q') }

@@

\Rule[leftskip=.2cm,rightskip=.1cm]{Q:Return}

@@

{ B; (\Gamma_R,Q_R); (\Gamma_R[\Vret/x], Q) \vdash \code{return}~x \dashv (\Gamma',Q') }

@@

\and \Rule[leftskip=.1cm,rightskip=.2cm]{Q:Update}

@@

We are only aware of two verified quantitative analysis systems.
Albert et al.~\cite{AlbertBGHR12} rely on the KeY tool to
automatically verify previously inferred loop invariants, size
relations, and ranking functions for Java Card programs.  However,
they do not have a formal cost semantics and do not prove the bounds
correct with respect to a cost model.  Blazy et al.~\cite{Blazy13}
have verified a loop bound analysis for CompCert's RTL intermediate
language.  However, there is no way to interactively derive bounds or
to deal with resources like memory usage.  Furthermore, Blazy et al.'s
automatic bound analysis does not compute symbolic bounds.
